\bf Model Visualization Techniques for a Social Network Model

### 2018-09-28 paper-child.Rnw
edit section 2

\noindent \textbf{Example Data:} To guide our visual exploration of SAOMs, we use two example data sources. The first is a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage. These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make node-link diagrams easier to see and to make any changes in the network more noticeable. %To determine which subset to select, we visualized all waves of the full network using the adjacency matrix visualization approach, which we show in Figure~\ref{fig:adjmatviz}. This adjacency matrix visualization is different from the one in Figure~\ref{fig:toyplots} because it does not show the node IDs on the axes. The network in Figure~\ref{fig:adjmatviz} is much larger than our toy data example, so we remove the node labels to remove clutter. In all three adjacency matrices, the ego and the alter nodes are ordered by node ID, 1-50, which were determined arbitrarily by the relationships in wave 1. 
%In Figure~\ref{fig:adjmatviz}, we show the adjacency matrix visualization of all three observations of this network, with the subset we selected for analysis outlined in red. This subset contains 16 actors, their ties and their drinking behavior for three timepoints. 
The node-link visualization of this data is provided in Figure~\ref{fig:alldataviz}. For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves. The actor-level categorical covariate, drinking behavior, has four values: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, and (4) drinks once a week. In Figure~\ref{fig:alldataviz}, the nodes are colored according to the drinking behavior of that student. Over time, we can see that the students tend to drink more and become increasingly isolated into smaller groups. An analysis of this type of data with a SAOM should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and behavioral forces at play.

The second data example we use is a collaboration network in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses. These sessions of congress correspond to the years of Barack Obama's presidency, from 2009-2016. (Details of how this data can be downloaded are provided by François Briatte at \url{https://github.com/briatte/congress}). In this network, ties are directed from senator $i$ to senator $j$ when senator $i$ cosponsors the bill that senator $j$ authored. There are hundreds of ties between senators when they are connected in this way, so we simplify the network by computing a single value for each pair of senators called the \textit{weighted propensity to cosponsor} (WPC). This value is defined in \citet{senate} as
\begin{equation}\label{eq:sen1}
WPC_{ij} = \dfrac{\sum\limits_{k=1}^{n_j} \frac{Y_{ij(k)}}{c_{j(k)}}}{\sum\limits_{k=1}^{n_j} \frac{1}{c_{j(k)}}}
\end{equation}
where $n_j$ is the number of bills in a congressional session authored by senator $j$, $c_{j(k)}$ is the number of cosponsors on senator $j$'s $k^{th}$ bill, where $k \in \{1,\dots, n_j\}$, and $Y_{ij(k)}$ is a binary variable that is 1 if senator $i$ cosponsored senator $j$'s $k^{th}$ bill, and is 0 otherwise. This measure ranges in value from 0 to 1, where $WPC_{ij} = 1$ if senator $i$ is a cosponsor on every one of senator $j$'s bills and $WPC_{ij} = 0$ if senator $i$ is never a cosponsor any of senator $j$'s bills.

@@

Because we require binary edges for SAOMs, we focus only on very strong collaborations, where $WPC$ is high. For a senate collaboration network $x$, edges are defined as
\begin{equation}\label{eq:wpc2}
  x_{ij} =
  \begin{cases}
                                   1 &  \text{if } WPC_{ij} > 0.25 \\
                                   0 & \text{if } WPC_{ij} \leq 0.25.
  \end{cases}
\end{equation}
The node-link diagram for the four senates during the Obama administration are shown in Figure~\ref{fig:senateplot}. We can see that Senate 111 is much more densely connected than the other three, which have one large, sparsely connected component with a few smaller connected groups throughout. Some prominent names are shown to demonstrate their place in the network.  \\

\noindent \textbf{Models:} To our example data, we fit three different SAOMs. Each SAOM uses a simple rate function, $\alpha_m$, and an objective function with two or three parameters. The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$: 
\begin{equation}\label{eq:m1}
f_i(x)^{M1} = \beta_1s_{i1}(x) + \beta_2s_{i2}(x),
\end{equation}
where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$ (see Table~\ref{tab:effects}. The second and third models, M2 and M3, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}. The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
\begin{align}\label{eq:m2m3}
f_i(x,z)^{M2} & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_3s_{i3}(x, z) \\
f_i(x)^{M3} & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_4s_{i4}(x),
\end{align}

@@

\noindent where $s_{i3}(x,z) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4}(x) =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$. These statistics are known as the \emph{number of jumping transitive triplets} (JTT) and the \emph{number of doubly achieved distances two effect} (N22), respectively. The first statistic emphasizes triad relationships formed between actors with different covariate values, while $s_{i4}$ emphasizes indirect ties between actors. The covariate for the friendship data is the drinking behavior, and is the number of bills authored for the senate data. These additional effects are shown in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, where the ties affected by the parameter value are represented by dotted lines.

It is ofen difficult to know for certain how to interpret a fitted value of a parameter. We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible. By graphically exploring these models, we aim to understand their effects, their interpretations, and the model fitting process better.

In this paper, we use three different model visualization techniques on SAOMs. First, Section~\ref{sec:minds} places the models in the data space. Then, Section~\ref{sec:collectmodels} explores visualizations of various collections of models. Finally, in Section~\ref{sec:algorithms} we use model visualization to explore the underlying algorithms in the SAOM fitting process. We conclude with a discussion in Section~\ref{sec:discussion}.
 %In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first. %Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
%In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:

@@

Another way to view a SAOM in the data space is with \textit{difference networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}. We created these networks by computing the difference between the adjacency matrices for the expected network and wave 1, and for wave 2 and the expected network. The lack of model fit becomes even more clear in Figure~\ref{fig:summary2}, with the XX NEED TO GO BACK AND REDO LINETYPES green XX edges representing ties that only show up in the expected network, and the red edges representing ties in the data that the model misses completely.  Again, the widths of the edges demonstrate the frequency of the ties in the simulations. On the left in Figure~\ref{fig:summary2}, we see the difference between the first wave of data and the expected network. We only see green and black edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that are not in the first wave of data. The additional green edges are all reciprocating black edges, meaning that the model predicts more reciprocity in ties than there actually is. On the right in Figure~\ref{fig:summary2}, we see the difference between the second wave of data and the expected network. As the model is simulating the second wave of data, we would hope to see few differences between the expected network and wave 2. However, there are six red edges, which are edges in the data that do not appear at all in the expected network, and there are several green edges that are in the expected network, but not in the data. Although there are some edges that appear in both the expected network and wave 2, the majority of edges in the expected network are not in wave 2, and there are several missing edges. We conclude that the model M1 is a poor fit to our example data, which have seen very clearly by viewing the model in the data space.

>>>>

\noindent \textbf{Example Data:} To guide our visual exploration of SAOMs, we use two example data sources. The first is a subset of the 50 actor dataset from the ``Teenage Friends and Lifestyle Study" that is provided on the \texttt{RSiena} webpage. These data come from \citet{friendsdata}, and we chose to only work with a subset of the data to make node-link diagrams easier to see and to make any changes in the network more noticeable. %To determine which subset to select, we visualized all waves of the full network using the adjacency matrix visualization approach, which we show in Figure~\ref{fig:adjmatviz}. This adjacency matrix visualization is different from the one in Figure~\ref{fig:toyplots} because it does not show the node IDs on the axes. The network in Figure~\ref{fig:adjmatviz} is much larger than our toy data example, so we remove the node labels to remove clutter. In all three adjacency matrices, the ego and the alter nodes are ordered by node ID, 1-50, which were determined arbitrarily by the relationships in wave 1. 
%In Figure~\ref{fig:adjmatviz}, we show the adjacency matrix visualization of all three observations of this network, with the subset we selected for analysis outlined in red. This subset contains 16 actors, their ties and their drinking behavior for three timepoints. 
The node-link visualization of this data is provided in Figure~\ref{fig:alldataviz}. %For model fitting, we condition on wave 1 and estimate the parameters of the models for the second and third waves. 
The actor-level categorical covariate, drinking behavior, has four values: (1) does not drink, (2) drinks once or twice a year, (3) drinks once a month, and (4) drinks once a week. In Figure~\ref{fig:alldataviz}, the nodes are colored according to the drinking behavior of that student. We can see that, as time passes, the students seem to drink more and become increasingly isolated into smaller groups. An analysis of this type of data with a SAOM should capture these dynamics in a way that allows the researcher to draw conclusions about the nature of the network and behavioral forces at play.

The second data example we use is a much more recent dataset on collaboration in the United States Senate during the $111^{th}$ through $114^{th}$ Congresses. These sessions of congress correspond to the years of Barack Obama's presidency, from 2009-2016. (Details of how this data can be downloaded are provided by François Briatte at \url{https://github.com/briatte/congress}). In this network, ties are directed from senator $i$ to senator $j$ when senator $i$ cosponsors the bill that senator $j$ authored. There are hundreds of ties between senators when they are connected in this way, so we simplify the network by computing a single value for each pair of senators called the \textit{weighted propensity to cosponsor} (WPC). This value is defined in \citet{senate} as
\begin{equation}\label{eq:sen1}
WPC_{ij} = \dfrac{\sum\limits_{k=1}^{n_j} \frac{Y_{ij(k)}}{c_{j(k)}}}{\sum\limits_{k=1}^{n_j} \frac{1}{c_{j(k)}}}
\end{equation}
where $n_j$ is the number of bills in a congressional session authored by senator $j$, $c_{j(k)}$ is the number of cosponsors on senator $j$'s $k^{th}$ bill, where $k \in \{1,\dots, n_j\}$, and $Y_{ij(k)}$ is a binary variable that is 1 if senator $i$ cosponsored senator $j$'s $k^{th}$ bill, and is 0 otherwise. This measure ranges in value from 0 to 1, where $WPC_{ij} = 1$ if senator $i$ is a cosponsor on every one of senator $j$'s bills and $WPC_{ij} = 0$ if senator $i$ is never a cosponsor any of senator $j$'s bills.

@@

Because we require binary edges for SAOMs, we focus only on very strong collaborations, where $WPC$ is high. For a senate collaboration network $x$, edges are defined as
\begin{equation}\label{eq:wpc2}
  x_{ij} =
  \begin{cases}
                                   1 &  \text{if } WPC_{ij} > 0.25 \\
                                   0 & \text{if } WPC_{ij} \leq 0.25.
  \end{cases}
\end{equation}
The node-link diagram for the four senates during the Obama administration are shown in Figure~\ref{fig:senateplot}. We can see that Senate 111 is much more densely connected than the other three, which have one large, sparsely connected component with a few smaller connected groups throughout. Some prominent names are shown to demonstrate their place in the network.  As with the friendship example, analyzing the senate data with a SAOM should capture the underlying dynamics in a way that gives insight into the impact the behavioral factors have on the senate's collaboration structure. By using data sets from very different contexts, we hope to assess possible differences in model performance. \\

\noindent \textbf{Models:} To our example data, we fit three different SAOMs. Each SAOM uses a simple rate function, $\alpha_m$, and an objective function with two or three parameters. The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$: 
\begin{equation}\label{eq:m1}
f_i(x)^{M1} = \beta_1s_{i1}(x) + \beta_2s_{i2}(x),
\end{equation}
where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$ (see Table~\ref{tab:effects}). The second and third models, M2 and M3, contain one additional parameter each in the objective function. We chose the additional parameters because they were significant ($p$-value $<$ 0.05) according to a Wald-type test provided in the \texttt{RSiena} software \citep{RSienaTest}. The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
\begin{align}\label{eq:m2m3}
f_i(x,z)^{M2} & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_3s_{i3}(x, z) \\
f_i(x)^{M3} & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_4s_{i4}(x),
\end{align}

@@

\noindent where $s_{i3}(x,z) = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4}(x) =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$. These statistics are known as the \emph{number of jumping transitive triplets} (JTT) and the \emph{number of doubly achieved distances two effect} (N22), respectively. The JTT effect emphasizes triads formed between actors with different covariate values, while the N22 effect emphasizes indirect ties between actors. The covariate for the friendship data is the drinking behavior, and the number of bills authored for the senate data. These additional effects are visualized in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, where the ties affected by the parameter value are represented by dotted lines. If the estimates of these parameters $\hat{\beta}_1, \dots, \hat{\beta}_4$ are positive, the model encourages the tie represented by the dotted line to form, while the opposite is true if the parameter estimate is negative.

%It is ofen difficult to know for certain how to interpret a fitted value of a parameter. We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible. By graphically exploring these models, we aim to understand their effects, their interpretations, and the model fitting process better.
% XXXX MOVE BELOW PAR TO SECTION 1 XXX
% In this paper, we use three different model visualization techniques on SAOMs. First, Section~\ref{sec:minds} places the models in the data space. Then, Section~\ref{sec:collectmodels} explores visualizations of various collections of models. Finally, in Section~\ref{sec:algorithms} we use model visualization to explore the underlying algorithms in the SAOM fitting process. We conclude with a discussion in Section~\ref{sec:discussion}.
 %In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first. %Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
%In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:

@@

Another way to view a SAOM in the data space is with \textit{difference networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}. We created these networks by computing the difference between the adjacency matrices for the expected network and wave 1, and for wave 2 and the expected network. The lack of model fit becomes even more clear in Figure~\ref{fig:summary2}, with the XX NEED TO GO BACK AND REDO LINETYPES green XX edges representing ties that only show up in the expected network, and the red edges representing ties in the data that the model misses completely.  Again, the widths of the edges demonstrate the frequency of the ties in the simulations. On the left in Figure~\ref{fig:summary2}, we see the difference between the first wave of data and the expected network. We only see green and black edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that are not in the first wave of data. The additional green edges are all reciprocating black edges, meaning that the model predicts more reciprocity in ties than there actually is. On the right in Figure~\ref{fig:summary2}, we see the difference between the second wave of data and the expected network. As the model is simulating the second wave of data, we would hope to see few differences between the expected network and wave 2. However, there are six red edges, which are edges in the data that do not appear at all in the expected network, and there are several green edges that are in the expected network, but not in the data. Although there are some edges that appear in both the expected network and wave 2, the majority of edges in the expected network are not in wave 2, and there are several missing edges. We conclude that the model M1 is a poor fit to our example data, which we have seen very clearly by viewing the model in the data space.
